```{r}
#install the necessary packages
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("randomForest")
#install.packages("forcats")
#install.packages("kableExtra")
#install.packages("caret")
#install.packages("e1071")
#install.packages("rpart")
#install.packages("corrplot")
#install.packages("ranger")
```
 

```{r}
#load the libraries
library("tidyverse")
library("ggplot2")
library("randomForest")
library("forcats")
library("kableExtra")
library("caret") # to calculate linear model performance
library("e1071") # Load library for SVR
library("rpart") # Load the rpart package for decision tree
library("corrplot")
library("ranger")
```



Data Preparation

1. Data Loading & Initial Inspection
```{r}
#get the working directory, and view the first rows of the dataset

getwd()
setwd("C:/Users/welcome/Desktop/Data Mining and Machine Learning")
housing_data <- read.csv("Housing Data_Same Region.csv")

head(housing_data)
```


2. Data Quality Checks

```{r}
#view the structure of the dataset
str(housing_data)
```

```{r}
#view the summary of the dataset
summary(housing_data)
```


```{r}
#check for missing values in the housing data
colSums(is.na(housing_data))
```

```{r}
#check for duplicates, using the unique identifier
sum(duplicated(housing_data$PARCELNO))
```

```{r}

# Get all duplicate parcels (with all columns)
duplicates <- housing_data %>% 
  group_by(PARCELNO) %>% 
  filter(n() > 1) %>%  # Keep only parcels appearing >1 time
  arrange(PARCELNO)    # Sort for easy viewing

# View first 6 duplicates
head(duplicates)

```


3. Feature Engineering & Type Conversion
```{r}

# Drop the first 3 columns
housing_data <- housing_data[, -c(1, 2, 3)]  


# Show new structure
head(housing_data)

```

```{r}

#move SALE_PRC as the target variable, to the last column.
housing_data <- relocate(housing_data, SALE_PRC, .after = last_col())

head(housing_data)
```


```{r}
#check for Null Values
colSums(is.na(housing_data))

```


```{r}

#check the dimension of the dataframe
dim(housing_data)

```

```{r}

# Convert avno60plus to a factor, since it is a binary operator
housing_data$avno60plus <- as.factor(housing_data$avno60plus)

head(housing_data) # Display the first few rows of the dataset
```


```{r}

# Convert month_sold to a factor
housing_data$month_sold <- as.factor(housing_data$month_sold)

head(housing_data)
```


```{r}
# Verify the change. Check structure to confirm avno60plus and month_sold, are now factor
str(housing_data)
```


```{r}
# Check outliers for the non-categorical variables
outliers <- boxplot(housing_data$SALE_PRC,
                   housing_data$LND_SQFOOT,
                   housing_data$TOT_LVG_AREA,
                   housing_data$SPEC_FEAT_VAL,
                   housing_data$RAIL_DIST,
                   housing_data$OCEAN_DIST,
                   housing_data$WATER_DIST,
                   housing_data$CNTR_DIST,
                   housing_data$SUBCNTR_DI,
                   housing_data$HWY_DIST,
                   housing_data$age,
                   names = c("PRICE", "SQFOOT", "LVG_AREA",
                           "SPEC_FEAT", "RAIL_DIST", "OCEAN_DIST",
                           "WATER_DIST", "CNTR_DIST", "SUBCNTR_DI",
                           "HWY_DIST", "age"),
                   main = "Outliers in Housing Data",
                   las = 2,          # Vertical labels
                   cex.axis = 0.7)   # Label size
```


```{r}

# Show total outlier counts for each variable
sapply(housing_data[sapply(housing_data, is.numeric)], 
       function(x) length(boxplot(x, plot = FALSE)$out))

```


```{r}

# Scatter plot to view outliers for SPEC_FEAT_VAL and Price (the highest outlier count)
ggplot(housing_data, aes(SALE_PRC, SPEC_FEAT_VAL, color=factor(structure_quality))) +
  geom_point() +
  labs(title = "SPEC_FEAT_VAL vs Price (Colored by Quality)")
```


Correlation Analysis
```{r}
#Correlation matrix for feature engineering
cor_matrix <- cor(housing_data[, sapply(housing_data, is.numeric)])
corrplot(cor_matrix,
method = "color",
type = "upper",
tl.col = "black",
addCoef.col = "black",
number.cex = 0.7)
```


Top correlations with SALE_PRC
```{r}
top_correlations <- cor_matrix[, "SALE_PRC"] %>%
sort(decreasing = TRUE) %>%
head(10)

kable(top_correlations, col.names = "Correlation with Sale Price") %>%
kable_styling()
```


```{r}
# Features sorted by correlation strength with Sale Price
df <- data.frame(Feature = names(top_correlations), 
                 Correlation = as.numeric(top_correlations))

# Create correlation plot with value labels
ggplot(df, aes(reorder(Feature, Correlation), Correlation, fill = Correlation > 0)) +
  geom_col() +
  geom_text(aes(label = round(Correlation, 2)),  # Add this line for numbers
            hjust = ifelse(df$Corr > 0, -0.1, 1.1),  # Automatic positioning
            size = 2.8) +  # Adjust text size
  coord_flip() +
  scale_fill_manual(values = c("red3", "blue3")) +  # Color coding
  theme_minimal() +
  theme(legend.position = "none") + # Hide fill legend
  labs(title = "Features Correlations with Sale Price")
```


```{r}
# Final structure check before modeling
str(housing_data)
summary(housing_data)
colSums(is.na(housing_data))
```


Splitting the data into training and testing data
```{r}
#splitting the data into training (70%) and testing (30%) sets
set.seed(123)

sample_size <- floor(0.7 * nrow(housing_data))
train_index <- sample(seq_len(nrow(housing_data)), size = sample_size)

#Create explicit train and test sets
train_data <- housing_data[train_index, ]
test_data  <- housing_data[-train_index, ]

head(train_data)
head(test_data)
```


Training set size: 9,752 observations  
Testing set size: 4,180 observations

Each set contains 14 variables, including the target variable `SALE_PRC`.  
This split ensures sufficient data for both model training and unbiased performance evaluation.
```{r}

dim(train_data)
dim(test_data)
```


Note: The month_sold variable is treated as a categorical feature. The model creates 11 dummy variables (February to December), using January as the reference month. Most months show no statistically significant effect on price, except May (month_sold5).

Properties located in areas with high airplane noise (avno60plus == 1) are expected to sell for about $65,090 less on average than those in quieter areas — and this difference is statistically significant.
```{r}
# Fit the linear regression model
lm_model <- lm(SALE_PRC ~ ., data = train_data)

# View model summary
summary(lm_model)
```


```{r}

# Predict on test data
lm_predictions <- predict(lm_model, newdata = test_data)

# Evaluate performance
lm_performance <- postResample(pred = lm_predictions, obs = test_data$SALE_PRC)

print(lm_performance)
```


A Support Vector Regression model with a linear kernel is fitted using the training dataset. The model is then used to predict sale prices on the test set, and its performance is evaluated using RMSE, MAE, and R-squared.
```{r}

# Train the SVR model with a linear kernel
svr_linear <- svm(SALE_PRC ~ ., data = train_data, kernel = "linear")

svr_linear_predictions <- predict(svr_linear, newdata = test_data) # Predict on test set

svr_performance <- postResample(pred = svr_linear_predictions, obs = test_data$SALE_PRC) # Evaluate performance using caret


print(svr_performance)

```

The RBF kernel, which is the default for Support Vector Regression, is used to capture non-linear relationships between predictors and sale price. This model is evaluated using standard performance metrics.
```{r}

# Train SVR model with the default RBF kernel
svr_rbf <- svm(SALE_PRC ~ ., data = train_data)  # RBF is the default kernel

svr_rbf_predictions <- predict(svr_rbf, newdata = test_data) # Predict on test data

svr_rbf_performance <- postResample(pred = svr_rbf_predictions, obs = test_data$SALE_PRC) # Evaluate performance


print(svr_rbf_performance)

```


```{r}

# Train SVR model with polynomial kernel
svr_poly <- svm(SALE_PRC ~ ., data = train_data, kernel = "poly")

svr_poly_predictions <- predict(svr_poly, newdata = test_data) # Predict on test data

svr_poly_performance <- postResample(pred = svr_poly_predictions, obs = test_data$SALE_PRC) # Evaluate performance


print(svr_poly_performance)

```


```{r}

# Fit the decision tree model
dectree_model <- rpart(SALE_PRC ~ ., data = train_data)

dectree_predictions <- predict(dectree_model, newdata = test_data) # Predict on the test set

dectree_performance <- postResample(pred = dectree_predictions, obs = test_data$SALE_PRC) # Evaluate the model


print(dectree_performance)

```


```{r}

# Fit the Random Forest model with 100 trees
rf_model_n100 <- randomForest(SALE_PRC ~ ., data = train_data, ntree = 100)

rf_predictions_n100 <- predict(rf_model_n100, newdata = test_data) # Predict on the test data

rf_performance_n100 <- postResample(pred = rf_predictions_n100, obs = test_data$SALE_PRC) # Evaluate performance


print(rf_performance_n100)

```


```{r}

# Fit the Random Forest model with 200 trees
rf_model_n200 <- randomForest(SALE_PRC ~ ., data = train_data, ntree = 200)

rf_predictions_n200 <- predict(rf_model_n200, newdata = test_data) # Predict on the test data

rf_performance_n200 <- postResample(pred = rf_predictions_n200, obs = test_data$SALE_PRC) # Evaluate performance


print(rf_performance_n200)
```


```{r}

# Fit the Random Forest model with 500 trees
rf_model_n500 <- randomForest(SALE_PRC ~ ., data = train_data, ntree = 500)

rf_predictions_n500 <- predict(rf_model_n500, newdata = test_data) # Predict on the test data

rf_performance_n500 <- postResample(pred = rf_predictions_n500, obs = test_data$SALE_PRC) # Evaluate performance


print(rf_performance_n500)
```


```{r}
#create a dataframe to put the RMSE of all the models
models_dataframe <- data.frame(
  Model = c("Linear Model", "SVR", "SVR (RBF)", "SVR (Poly)", 
            "Decision Tree", "RF (n=100)", "RF (n=200)", "RF (n=500)"),
  RMSE = c(
    lm_performance[1],           
    svr_performance[1],      
    svr_rbf_performance[1],
    svr_poly_performance[1],
    dectree_performance[1],
    rf_performance_n100[1],
    rf_performance_n200[1],
    rf_performance_n500[1]
  )
)

# Print the cleaned dataframe
print(models_dataframe)
```


```{r}

ggplot(models_dataframe, aes(x = Model, y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +  # Create bars for RMSE
  labs(title = "RMSE Comparison for Different Models", 
       x = "Model", 
       y = "RMSE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Find the row with the minimum RMSE
best_model <- models_dataframe[which.min(models_dataframe$RMSE), ]

# Print the result
print(best_model)
```


```{r}

# After identifying best_model
print(paste("Tuning parameters for:", best_model$Model))

# Use the same tree count from the best model (n=100)
num_trees <- 100  # Directly using the known best model specification

# Proceed with the tuning :
control <- trainControl(method = "cv", number = 5)

tune_grid <- expand.grid(
  mtry = 2:7,
  min.node.size = c(1, 5, 10),
  splitrule = "variance"
)

set.seed(123)
rf_tuned_ranger <- train(
  SALE_PRC ~ .,
  data = train_data,
  method = "ranger",
  metric = "RMSE",
  tuneGrid = tune_grid,
  trControl = control,
  num.trees = num_trees  # references the best model's tree count
)

# View best parameters and performance
print(rf_tuned_ranger)

```

After performing hyperparameter tuning on the Random Forest model using the ranger package and 5-fold cross-validation, the optimal parameters identified were mtry = 7 and min.node.size = 1. The final model was rebuilt using these values and evaluated on the test data, showing improved predictive performance.
```{r}
# Rebuild the final tuned model using ranger with best hyperparameters
set.seed(123)
final_rf_model <- ranger(
  formula = SALE_PRC ~ .,
  data = train_data,
  num.trees = 100,
  mtry = 7,
  min.node.size = 1,
  splitrule = "variance"
  )

# Predict on the test set
final_predictions <- predict(final_rf_model, data = test_data)$predictions

# Evaluate the model and print results
final_performance <- postResample(pred = final_predictions, obs = test_data$SALE_PRC)

print(paste("Final RMSE:", round(final_performance["RMSE"], 2)))
print(paste("Final R-squared:", round(final_performance["Rsquared"], 4)))
print(paste("Final MAE:", round(final_performance["MAE"], 2)))

print(paste("Number of trees:", final_rf_model$num.trees))
print(paste("mtry (features per split):", final_rf_model$mtry))
print(paste("Minimum node size:", final_rf_model$min.node.size))
print(paste("Split rule:", final_rf_model$splitrule))

```


The performance of all models was compared using Root Mean Squared Error (RMSE). The Linear Regression and Decision Tree models underperformed, while Support Vector Regression (SVR) with RBF and Polynomial kernels showed improvement. However, the Random Forest model consistently yielded the best results.

After performing hyperparameter tuning using a grid search on mtry and min.node.size, the Tuned Random Forest model achieved the lowest RMSE (104,397.3), outperforming the previously best RF (n=100) model. This confirms the effectiveness of tuning in improving predictive accuracy.
```{r}
# Create a dataframe to compare model performance by RMSE
models_dataframe_2 <- data.frame(
  Model = c("Linear Model", "SVR", "SVR (RBF)", "SVR (Poly)", 
            "Decision Tree", "RF (n=100)", "RF (n=200)", "RF (n=500)", "Tuned RF"),
  
  RMSE = c(
    lm_performance[1],           
    svr_performance[1],      
    svr_rbf_performance[1],
    svr_poly_performance[1],
    dectree_performance[1],
    rf_performance_n100[1],
    rf_performance_n200[1],
    rf_performance_n500[1],
    final_performance["RMSE"]
  )
)

# Print the comparison table
print(models_dataframe_2)

```


```{r}
ggplot(models_dataframe, aes(x = reorder(Model, RMSE), y = RMSE, fill = Model)) +
  geom_bar(stat = "identity") +  
  coord_flip() +
  labs(x = "", y = "RMSE") +
  theme_minimal()
```


```{r}
# Confirm final model parameters
print(paste("Final RMSE:", final_performance["RMSE"]))
print(paste("mtry:", final_rf_model$mtry))
print(paste("Node size:", final_rf_model$min.node.size))
```


```{r}

# Save ONLY the tuned model (contains all parameters + learned structure)
saveRDS(final_rf_model, "final_tuned_rf_model.rds")

```


```{r}
# Load the saved tuned model
tuned_model <- readRDS("final_tuned_rf_model.rds")
tuned_model

```


```{r}
# Create data with given input
new_property <- data.frame(
  PARCELNO = 728980145245,
  LND_SQFOOT = 11247,
  TOT_LVG_AREA = 4552,          
  SPEC_FEAT_VAL = 2105,         
  RAIL_DIST = 4871.9,
  OCEAN_DIST = 18507.2,
  WATER_DIST = 375.8,
  CNTR_DIST = 43897.9,
  SUBCNTR_DI = 40115.7,
  HWY_DIST = 41917.1,
  age = 42,
  avno60plus = 0,
  structure_quality = 5,
  month_sold = 8
)

# Convert factors with ALL levels the model expects
new_property$avno60plus <- factor(new_property$avno60plus, levels = 0:1)  # Both levels
new_property$month_sold <- factor(new_property$month_sold, levels = 1:12) # All months

# Predict
predicted_price <- predict(tuned_model, data = new_property)$predictions

# Format output
print(paste("Predicted Price: $", 
            format(round(predicted_price, 0), big.mark = ","), 
            sep = ""))
```


```{r}
# Calculate metrics - Predict on the test set
test_predictions <- predict(tuned_model,      
                            data = test_data)$predictions

#Evaluate RMSE, R‑squared, MAE -------------------------------
test_metrics <- postResample(pred = test_predictions,
                             obs  = test_data$SALE_PRC)

print(paste("Test RMSE:", round(test_metrics["RMSE"], 0)))  # Typical error margin
print(paste("Test R²:", round(test_metrics["Rsquared"], 3))) # Variance explained
```


```{r}
# Find comparable properties in the training data
similar <- train_data[
  abs(train_data$TOT_LVG_AREA - 4552) < 500 &       # ±500 sqft
  train_data$structure_quality == 5 &               # Same quality
  abs(train_data$age - 42) < 5,                    # ±5 years age
  c("SALE_PRC", "TOT_LVG_AREA", "structure_quality", "age")]

# Calculate price per sqft
similar$price_per_sqft <- similar$SALE_PRC / similar$TOT_LVG_AREA
summary(similar$price_per_sqft)

# Your prediction's price/sqft
your_pps <- 1277657 / 4552  # ≈ $281/sqft

# Compare to market
cat("Your price/sqft:", round(your_pps, 0), "\n",
    "Neighborhood median:", median(similar$price_per_sqft, na.rm=TRUE))
```

